The markings of the trade: exploring the patterns of olive oil production in Roman Baetica

### 2020-02-07 Paper/jaccardbaetica.tex
new changes intro final

% hipotesis

% Debate

%más específico--- olive oil
%incluir algo que enlace con roman province

\xavi{no digo nada porque entiendo que es una primera versión de intro no?}

\section{The Amphoric production in Baetica province}

\xavi{y si esta sección detalla directamente la producción amfórica de Dressel 20 en el Guadalquivir?}

@@

The majority of olive oil amphorae produced in this province and shipped through the provinces belong to Dressel 20 type \citep{dressel,martin-kilcher_romischen_1994}. This amphora is commonly associated with the transportation of Baetican olive oil during the Roman Empire \citep{berni_millet_epigrafianforica_2008}. Most Dressel 20 were marked in stamps, inked in \textit{tituli picti} and incise in \textit{graffiti} with different information. However, inscriptions and \textit{tituli picti} have not enough analysed due to the fragmentation of the material or the shortage of samples \citep{aguilera_evolucion_2007,rovira_guardiola_grafitos_2007}. By contrast, stamps are the most studied in this type of amphorae.
The fascination aroused by stamps has caused it to be one of the inscriptions with the most publications about its origin and meaning. Its endure in time become it an essential element that were used to mark amphorae during almost three centuries \citep{remesal_sellar_2016}. Frequently, they were marked mainly in handles but seldom in rims and body \citep{millet_anforas_1998}. 
The information of the stamps is shown in different forms and letter content and it seems that there was not a unique criterion. Stamps are mostly formed by a code of three letters and they can appear in a abbreviated form or complete and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}.

@@

%hipotesis: si esto es así entonces es asá
Here, therefore, this study proposes an evolutionary approach to explore the effect of Baetican olive oil production by computing the distribution of stamps. To do that, three case study have been studied to analyse the relation between production centres (Baetica) and consumption centres (Britannia and Germania).

The assumption of this study is analysing the effect of the production patterns between different centres.  This effect could affect due to different aspects: a) correlation between spatial distance and centres based on the idea that closer workshops concentrate similar amphoric stamp in a specific area than the distant workshops and b) groups of similar stamps were concentrated in a specific Roman province. In the case of Baetica province, we want to identify the role of the stamps in the organization of the workshop. In Roman provinces, our aim is detecting groups of stamps concentrated in a area or if some groups have an important role for the exportation of olive oil in those provinces.

%resumen del paper

%hablar de las tres áreas de producción: hispalis, corduba y astigi

>>>>

\subsection{The Amphoric production in Baetica province}

%\xavi{y si esta sección detalla directamente la producción amfórica de Dressel 20 en el Guadalquivir?}

@@

The majority of olive oil amphorae produced in this province and shipped through the provinces belong to Dressel 20 type \citep{dressel_ricerche_1878,
millet_anforas_1998}. This amphora is commonly associated with the transportation of Baetican olive oil during the Roman Empire \citep{berni_millet_epigrafianforica_2008}. Most Dressel 20 were marked in stamps, inked in \textit{tituli picti} and incise in \textit{graffiti} with different information. However, inscriptions and \textit{tituli picti} have not enough analysed due to the fragmentation of the material or the shortage of samples \citep{aguilera_evolucion_2007,rovira_guardiola_grafitos_2007}.

\subsection{A potential indicator of Roman economy}

Stamps are the most studied in this type of amphorae. The fascination aroused by stamps has caused it to be one of the inscriptions with the most publications about its origin and meaning. Its endure in time become it an essential element that were used to mark amphorae during almost three centuries \citep{remesal_sellar_2016}. Frequently, they were marked mainly in handles but seldom in rims and body \citep{millet_anforas_1998}. 
The information of the stamps is shown in different forms and letter content and it seems that there was not a unique criterion. Stamps are mostly formed by a code of three letters and they can appear in a abbreviated form or complete and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}.

@@

%hipotesis: si esto es así entonces es asá

The assumption of this study is analysing the effect of the production patterns between different centres. We are specially interested in identifying links between producers and consumption centres using amphoric stamps. To do that, we use the CEIPAC database to collect the stamps from different places. The CEIPAC dataset contains over 50.000 of epigraphy records found in amphorae, mostly from \textit{Monte Testaccio}.

Three case study have been studied in order to analyse the relation between production centres (Baetica) and consumption centres (Britannia and Germania). In the case of Baetica province, we want to identify the role of the stamps in the organisation of the workshop. In Roman provinces, our aim is detecting groups of stamps concentrated in a area or if some groups have an important role for the exportation of olive oil in those provinces. This economic connection could affect due to different aspects: a) correlation between spatial distance and centres based on the idea that closer workshops concentrate similar amphoric stamp in a specific area than the distant workshops and b) groups of similar stamps were concentrated in a specific Roman province.

Therefore, this study proposes an evolutionary baseline to explore the distribution of Baetican olive oil production by computing spatial correlation between stamps. A way to analyse is by using a quantitative framework to measure the similarity. Here we use a ecological approach based on three steps: a) to detect similarities between stamp codes, b) to explore a potential spatial correlation and c)to establish a correlation between similarity of stamp and spatial distance.
